---
title: "Homework for Advanced R programing - Class 7 - Script and report automation (RMarkdown)"
author: "Rafał Rysiejko"
date: "09/04/2020"
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
      
---

```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

The recorded music industry was worth $19.1 billion in 2018, which was almost a double-digit gain (9.7%) from the year prior [^1]. In this entertainment market, an increasingly large number of song products are introduced each year. However, only a small group achieves mainstream success, and among these, the distribution of market success is becoming increasingly skewed. By analyzing the top songs from the popular music streaming service provider, we can investigate if some underlying patterns make a song truly hit.  

This assignment will try to determine whether there are some distinctive features of song that make it successful and if so, are there any patterns among those features.
Dataset [^2] used for this assignment contains information about the top 50 most listened songs in the world on music streaming platform Spotify in 2019. Each song has a set of 13 variables:  

   
\setlength{\leftskip}{6cm}

Table 1: Variable description

\setlength{\leftskip}{0pt} 


```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis',fig.align = "center"}
tabl <- "
| Variable    | Description         |
|-----------|:------------------------| 
| Track.Name          |Song Title |
| Artist.Name         | Artist performing the song |
| Genre               | The genre of the track   |
| Beats.Per.Minute    | Variable describing the tempo of the song.  |
| Energy              | The energy of a song - the higher the value, the more energtic song.    |
| Danceability        | The higher the value, the easier it is to dance to this song.    |
| Loudness..dB.       | The higher the value, the louder the song.   |
| Liveness            | The higher the value, the more likely the song is a live recording.    |
| Valence             | The higher the value, the more positive mood for the song.  |
| Length              | The duration of the song. |
| Acousticness        | The higher the value the more acoustic the song is. |
| Speechiness         | The higher the value the more spoken word the song contains.  |
| Popularity          | The higher the value the more popular the song is. |
"
cat(tabl) 
```


## Exploratory data analysis

Installing and running the libraries:
```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
requiredPackages = c("tidyverse","factoextra","stats","clustertend","flexclust","ggforce"
,"fpc","cluster","ClusterR","knitr","kableExtra","DataExplorer","caret",
"reshape2","corrplot","labdsv","smacof","clusterSim","pastecs","psych","pca3d","pls") 
for(i in requiredPackages){if(!require(i,character.only = TRUE)) install.packages(i)} 
for(i in requiredPackages){library(i,character.only = TRUE) } 
```

Loading the data:
```{r echo=T}
data_full <- read.csv("top50.csv",stringsAsFactors = F)
data_full <- data_full[,2:14]
data_init <- data_full
```

```{r echo=F}
x1 <- t(round(stat.desc(data_init[,c(4:13)]),2))
x1 <- x1[,c(4:6,8:13)]

kable(x1,caption = "Summary statistics of the dataset")%>% 
kable_styling(latex_options="hold_position",bootstrap_options = c("striped", "hover"))
```

Songs in the data set are grouped into `r length(unique(data_init$Genre))` genres.
 
```{r echo=F,message=FALSE,out.width = '70%',fig.align = "center",fig.cap="\\label{fig:figs}Genre distribution."}
data_init %>% count(Genre) %>%
  ggplot(aes(reorder(Genre, n), n)) + geom_col(fill="cyan3") + coord_flip() + ylab("Count") +xlab("") + theme_minimal()
```
 
As there a lot of subgenres with a low number of instances, a new variable `aggregate.genres` is created.
  
  
```{r echo=T,message=FALSE}
data_init$aggregate.genre <- NA
data_init$aggregate.genre <- ifelse(data_init$Genre  %in% c("canadian pop","pop","dance pop",
                        "electropop","panamanian pop","pop house","australian pop"),"pop",
ifelse(data_init$Genre  %in% c("dfw rap","country rap","canadian hip hop","atl hip hop"),"hip hop",
ifelse(data_init$Genre  %in% c("r&b en espanol","latin"),"latin",
ifelse(data_init$Genre  %in% c("reggaeton flow","reggeaeton"),"reggaeton",
ifelse(data_init$Genre  %in% c("edm","trap music","brostep"),"electronic",
ifelse(data_init$Genre  %in% c("escape room","boy band","big room"),"other",data_init$Genre))))))

data_init$aggregate.genre <- as.factor(data_init$aggregate.genre)
``` 
 
  
  
```{r echo=F,message=FALSE,out.width = '70%',fig.align = "center",fig.cap="\\label{fig:figs} Aggregated genre distribution."}
data_init %>% count(aggregate.genre) %>%
  ggplot(aes(reorder(aggregate.genre, n), n)) + geom_col(fill="cyan3") + coord_flip() + ylab("Count") +xlab("") + theme_minimal()
```
 

  
```{r echo=F,message=FALSE,out.width = '80%',fig.align = "center",fig.cap="\\label{fig:figs} Histograms of numerical variables."}
gather(data_init[,c(4:13)]) %>% 
  ggplot(., aes(value)) + 
  geom_histogram(aes(y =..density..), 
                 col="black", 
                 fill="lightblue", 
                 alpha=.2) + 
  geom_density(col="cyan3")+
  facet_wrap(~key, scales = 'free') + theme_minimal()
```

Variables `Liveness` and `Speechiness.` are highly positively skewed, suggesting the existence of outliers.

```{r echo=T,message=FALSE}
transformed_var <- c("Liveness", "Speechiness.")
data_init <- data_init %>% mutate_at(vars(transformed_var), log)
``` 

Distribution of variables `Liveness` and `Speechiness.` after log transformation
  

To further investigate the relationship between variables, a correlation analysis was carried using a Pearson correlation coefficient.  


```{r echo=FALSE, message=FALSE,out.width = '50%',fig.align = "center",fig.cap="\\label{fig:figs} Correlation matrix."}

data_corr<- select_if(data_init, is.numeric)
corrplot::corrplot(cor(data_corr), type="lower", method="number",order ="alphabet")

```
  
  
From Figure 4, we can see a high positive correlation between variables `Speechiness` and `Beats.Per.Minute`, typical for rap songs. Another relationship is visible between variables `Energy` and `Loudness.dB.` suggesting that highly energetic songs tend to be louder. Also, a more energetic song, on average less acoustic.



[^1]: <https://www.forbes.com/sites/hughmcintyre/2019/04/02/the-global-music-industry-hits-19-billion-in-sales-in-2018-jumping-by-almost-10/#52c7da3e18a9>

[^2]: Dataset available at: <https://www.kaggle.com/leonardopena/top50spotify2019/data>
